---
title: "An introduction to R/RStudio"
subtitle: "Data structures, functions, packages and the RStudio environment"
format: html
---

Install packages

```{r, message=FALSE, warning=FALSE}}
install.packages("tidyverse", "dplyr")
library(tidyverse)
library(dplyr)
```

You can check your version of R using the following commands:

```{r}
R.version
```

## Basic R operations

We will now cover some basic operations and values within R.

We can perform basic mathematical operations:

```{r}
#| context: interactive

# Addition and Subtraction
3 + 2
3 - 2

# Multiplication and Division
3 * 2
3 / 2

# Exponents
3^2
2^3

# Constants
pi
exp(1)  # base of natural logarithm
```

In addition to numbers, variables can also take on other values:

```{r}
#| context: interactive

Inf
1 + Inf

# Missing Values
NA
1 + NA

# NULL Values
NULL
```

## Data types, classes and variables

There are many types of data in R, here are some commonly used:

Numeric - Decimal and 'whole' numbers (the most common numeric type)
```{r}

class(1.1)
```

Character - Text data in quotes
```{r}

class("hello world!")
class("42") # Numbers in quotes are characters
```

Logical - Boolean values for conditional logic
```{r}

class(TRUE)
class(F) # F is shorthand for FALSE
```

Factor - a data type for categorical variables with fixed levels (categories).

In the example below, we create a vector of letters, some of which are repeated. However, the levels within are limited to each individual letter.
```{r}

# Define a factor using letters
f <- factor(letters[c(1, 1, 2, 2, 3:10)])  # Create factor from letters with some repetition
f
```

```{r}

class(f)  # Check class
```

Storing and manipulating variables

We commonly assign numbers and data to variables, which we can then compute directly:
```{r}

# Define objects x and y with values of 3 and 2 respectively:
x <- 3
y <- 2

# Some calculations with the defined objects x and y:
x + y
x * y
```

R is case sensitive, so `X` and `x` are not the same object!
```{r}

X * y # This will throw an error
```

## Data structures

`R` offers several data structures that serve different purposes. Each structure is designed to handle specific types of data organization, from simple one-dimensional vectors to complex nested lists.

Vector

- One-dimensional sequence of elements
- All elements must be of the same type (numeric, character, etc.)
```{r}

v1 <- 1:12  # Create numeric vector using sequence
v2 <- c(2,4,1,5,1,6, 13:18)  # Create numeric vector by combining values and sequence
v3 <- c(rep('aa',4), rep('bb',4), rep('cc',4))  # Create character vector with repeated values

v1
v2
v3
```

Now let's check the class of each vector:
```{r}

class(v1)  # Check class of v1
class(v2)  # Check class of v2
class(v3)  # Check class of v3
```

Matrix and array

- Two-dimensional (matrix) and three-dimensional (array) arrangement of elements
- All elements must be of the same type
- Organized in rows and columns

```{r}

m1 <- matrix(v1, nrow=3, ncol=4)  # Create matrix filled by column
arr <- array(v1, dim=c(2,2,3))  # Create 3D array

m1
```

```{r}

arr
```

Let's check the class of the matrix and array:
```{r}

class(m1)  # Check class of matrix
class(arr)  # Check class of array
```

If you aren't sure which type of data structure you are working with, whether it is an array, matrix or vector, you can use the `is()` function to check. This will return `TRUE` or `FALSE` depending on what the structure is.

```{r}

# You can check in two ways:

# Like this:
is(m1, "matrix")  # Check if m1 is a matrix
is(m1, "array")   # Check if m1 is an array
is(v1, "vector") # Check if m1 is a vector

# Or like this:
is.matrix(arr)  # Check if arr is a matrix
is.array(arr)   # Check if arr is an array
is.vector(m1) # Check if m1 is a vector
```

You may have noticed that `m1` is a matrix and an array, whilst `arr` is an array but not a matrix. This is because a matrix is essentially a three-dimensional array with one dimension set to 1.

Data frame

- Two-dimensional structure similar to a spreadsheet
- Different columns can contain different types of data
- Most common structure for statistical analysis

Suppose we have a data frame of students' grades and demographics:
```{r}

# Create a data frame
students <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Gender = c("F", "M", "M"),
  Math = c(95, 88, 92),
  Science = c(89, 94, 90)
)

students
```

Check the class of the dataframe and a column, and display the structure of the dataframe:
```{r}

class(students)  # Check class of dataframe
is.data.frame(students)  # Check if students is a dataframe
class(students$Name)  # Check class of Name column
str(students)    # Display structure of dataframe
```

When working with dataframes, we often want to select specific columns. We can use the `$` operator to do this.
```{r}

students$Name  # print Name column
students$Gender  # print Gender column
```

Or we can use the `[[` operator:
```{r}

students[["Math"]]    # print Math column
students[["Science"]] # print Science column
```

Either is fine, but you may find the `$` operator is quicker and easier to use.

We can assign values within a specific column or row to a new variable:
```{r}

# Get the "Math" column
students_math <- students$Math
print(students_math)

# Get the first row
students_alice <- students[1, ]
print(students_alice)
```

But there are always lots of ways to do the same thing in `R`. For example, here are different ways by which we can extract all students who are male:

```r
students[students$Gender == "M", ]         # Basic subsetting
subset(students, Gender == "M")            # using the subset() function
students %>% filter(Gender == "M")         # using the filter() function from the dplyr package 
```

## Logical operators, control flow and functions

We commonly use logical operators in `R` to help make decisions in code and are essential in tasks like subsetting data, controlling loops, writing conditional statements, and filtering data.


| **Operator** | **Summary**                |
|--------------|----------------------------|
| `<`          | Less than                 |
| `>`          | Greater than              |
| `<=`         | Less than or equal to     |
| `>=`         | Greater than or equal to  |
| `==`         | Equal to                  |
| `!=`         | Not equal to              |
| `!x`         | NOT x                     |
| `x | y`      | x OR y                    |
| `x & y`      | x AND y                   |

Some examples using the `students` dataframe:
```{r}

# Less than (<) - Students with Math below 90
subset(students, Math < 90) # Returns Bob (88)

# Equal to (==) - Male students
subset(students, Gender == "M") # Returns Bob & Charlie

# Not equal to (!=) - Non-female students
subset(students, Gender != "F") # Returns Bob & Charlie

# OR (|) - Math OR Science > 90
subset(students, Math > 90 | Science > 90) # Returns all three students

# AND (&) - Math & Science > 90
subset(students, Math > 90 & Science >= 90) # Returns Charlie only
```

### Basic statistical functions in `R`

We can also perform basic statistics and operations on variables, such as getting the variance, standard deviation and summary statistics. 
You can do this using built-in functions in `R` including `var()`, `sd()`, `sum()`, `mean()`, `min()`, and `max()`.
```{r}

# Create a vector of 10 random numbers between 1-100 using the sample() command
x <- sample(1:100, 10)

# Variance and Standard Deviation
var(x)   # Calculates variance
sd(x)    # Calculates standard deviation

# Basic Statistics
sum(x)   # Sums all elements
mean(x)  # Calculates mean

# Range Functions
min(x)   # Finds minimum
max(x)   # Finds maximum
```

## Miscellaneous commands

Here are some other commands that will be useful when working with `R` more generally:

Directory and Workspace Management

The `setwd()` command sets the working directory, which is the folder where `R` looks for files to read and where it saves output files. You can check your current working directory using `getwd()`. It is important to set your working directory appropriately, otherwise, you may run into issues when trying to read or write files, or source functions from other scripts.
```{r}
getwd() # Get current working directory
# setwd("your/path/here")  Set working directory to your path
dir() # List files in current directory
```

Environment Management

The `ls()` command lists all objects in the current workspace. You can remove specific objects using `rm()`, or clear the entire workspace with `rm(list = ls())`. This is useful for cleaning up your environment before starting a new analysis.
```{r}
# List objects in workspace
ls()
```

```{r}
# Remove all objects from workspace
rm(list = ls())
```

## Packages

Packages are collections of functions, data sets, and documentation bundled together to extend the functionality of `R`.
They are not part of the base `R` installation but can be easily added and used in your environment.

`R` packages can:

- Add functions: They contain pre-written functions that simplify common tasks or complex analyses. For example, packages like `ggplot2` and `dplyr` offer powerful tools for data visualization and manipulation.

- Provide data: Some packages include data sets that can be used for testing or teaching purposes. For example, the `datasets` package provides a collection of sample data sets.

- Enable special features: Packages can implement specialized features like statistical models, machine learning algorithms, or tools for web scraping, reporting, and more.

### How to use packages in R

Installing: You can install a package from CRAN (the Comprehensive R Archive Network) using the `install.packages()` function.
```{r}
install.packages("ggplot2")
```

Loading: Once installed, you can load the package into your R session with the `library()` function.
```{r}
library('ggplot2')
```

Usage: After loading the package, you can use its functions. For example, with `ggplot2`, you can create a plot like this:
```{r}
#| context: interactive
ggplot(data = mtcars, aes(x = mpg, y = hp)) + 
  geom_point()
```

Packages are hosted on several repositories, CRAN being the most common. Other repositories include Bioconductor (for bioinformatics) and GitHub. The `install.packages()` function installs packages from CRAN, while for GitHub packages, you can use the `devtools` or `remotes` package to install directly from a GitHub repository.

Popular R packages include:

- `ggplot2`: A powerful package for data visualization based on the grammar of graphics.
- `dplyr`: A package for data manipulation (filtering, selecting, grouping, etc.).
- `tidyr`: Used for tidying data, such as reshaping and pivoting.
- `shiny`: For building interactive web applications in `R`.

We will be using `tidyverse` - a collection of packages for data manipulation and visualization including `dplyr`, `tidyr`, and `ggplot2` - in this workshop.

You can find and install R packages from a number of sources:

- [CRAN](https://cran.r-project.org/): The main repository for `R` packages.
- [Bioconductor](https://www.bioconductor.org/): A repository specializing in bioinformatics packages.
- [GitHub](https://github.com/): Many `R` developers host their packages on GitHub, which you can install using `devtools` or `remotes` packages.

## Data visualization using `ggplot2()`

One of the main benefits of `R` is to create publication quality figures and graphs. There are a number of different functions within `R` that we can use.

We will now briefly cover `ggplot2()` as it is the most versatile and used approach to create complex figures.

`ggplot2` is a powerful `R` package for creating complex and customizable data visualizations.**
It provides a systematic approach to building plots by combining two main components: geometries (`geom`) and aesthetics (`aes`).

`plot = geometric (points, lines, bars) + aesthetic (color, shape, size)`

Geometries (`geom`): These define the type of plot or visual elements you want to display. Common `geoms` include:

- `geom_point()`: Displays data points (scatter plot).
- `geom_line()`: Plots lines connecting data points (line plot).
- `geom_bar()`: Creates bar charts.
- `geom_histogram()`: Displays histograms using counts for continuous data.
- `geom_boxplot()`: Creates box plots.

Aesthetics (`aes`): These define how data is mapped to visual properties. The aesthetics determine the appearance of the plot, such as:

- `color`: Specifies the color of the points, lines, or bars.
- `shape`: Defines the shape of data points (e.g., circles, squares).
- `size`: Controls the size of the points or lines.

Importantly, `ggplot2()` is built upon the layering of different components. For example, you can simply add more `aes` components to add a line of best fit, and standard error:


Getting Help

You can access help documentation for functions and packages in `R` using the `?` or `??` commands. `?` is for direct help on a specific function, object, or topic when you know its exact name:
```{r}
# Run these one by one

# Get help on a specific function
?mean

# Help on datasets
?mtcars

# Help on packages
?dplyr
```

Conversely, `??` performs a broader search across all installed packages' documentation:
```{r}
# Run these one by one

# Search for topics related to "regression"
??regression

# Search for anything related to "anova"
??anova

# Search for cases of the word "bread" in the sandwich package
??sandwich::bread
```


Here are some shortcuts that you can use in RStudio:

| Shortcut | Action |
|----------|---------|
| `Ctrl + L` | Clean console |
| `Ctrl + Shift + N` | Create a new script |
| `↑` | Access command history |
| `Ctrl(hold) + ↑` | Search command history with current input |
| `Ctrl + Enter` | Execute selected code in script |

These shortcuts work on Windows/Linux. For Mac, replace Ctrl with Cmd (⌘).